---
title: "SF Housing"
author: "Thomas Lumley"
date: "17 May 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data source

These data come from [Eric Fischer's github](https://github.com/ericfischer/housing-inventory), and resulted from some fairly obsessive data collection described [here](https://experimental-geography.blogspot.co.nz/2016/05/employment-construction-and-cost-of-san.html)

My aim is to provide better graphs and some slightly different models

### Reading the data
Unfortunately, while most of the data file is comma-separated, the first two columns are separated only by a space, so we need a bit of gymnastics
```{r}
combined<-read.table("../combined",sep=",")
combined<-combined[,1:6]
year<-as.integer(gsub(" .*","",as.character(combined[,1])))
medrent<-as.numeric(gsub("^[0-9]+","",as.character(combined[,1])))
SF<-cbind(year=year,median_rent=medrent, combined[,-1])
names(SF)<-c("year","median_rent","housing_units","net_new_housing","employment","total_wages","CPI")
head(SF)
summary(SF)
```

## Trends

Nominal rental rates
```{r}
plot(median_rent~year, data=SF, type="p",log="y")
with(SF, lines(supsmu(year,median_rent),col="purple",lwd=2))
```

Real rental rates, after 1948, with a scatterplot smoother
```{r}
SF$real84_median_rent<-with(SF,median_rent/CPI)
plot(real84_median_rent~year, data=subset(SF, year>=1948), type="p",log="y",
     xlab="year",ylab="Median rent (1983 dollars)")
with(subset(SF,year>=1948), lines(supsmu(year,real84_median_rent),col="purple",lwd=2))
```

With trend line, getting rid of years with basically no data
```{r}
SFok<-subset(SF, year>=1948 & year <=2015)
SFok$real84_median_rent<-with(SFok,median_rent/CPI)
plot(real84_median_rent~year, data=SFok, type="p",log="y",
     xlab="year",ylab="Median rent (1983 dollars)")
with(SFok, lines(supsmu(year,real84_median_rent),col="sienna",lwd=2))
cat("annual increase=",
    exp(coef(trend_model<-glm(real84_median_rent~year,family=gaussian(log),data=SFok))[2]))
with(SFok, lines(year, fitted(trend_model),lty=2,col="blue"))
legend("bottomright",lty=c(1,2),col=c("sienna","blue"),
       bty="n",legend=c("smoother","constant real % increase"))
```

## Models

Eric Fischer used a multiplicative model for powers of explanatory variables, or more-or-less equivalently, a linear model for logarithms. Also, he uses `log10` rather than `log`
```{r}
pricemodel<-lm(log10(real84_median_rent)~log10(housing_units)+log10(employment)+log10(total_wages/CPI),
               data=SFok,na.action=na.exclude)
summary(pricemodel)
plot(real84_median_rent~year, data=SFok, type="p",log="y",
     xlab="year",ylab="Median rent (1983 dollars)")
lines(SFok$year, 10^(fitted(pricemodel)),col="purple")
```

Not quite what he found: maybe due to minimising squared error on the log scale rather than the raw scale?
```{r}
pricemodel2<-glm(real84_median_rent~log10(housing_units)+log10(employment)+log10(total_wages/CPI),
               data=SFok,na.action=na.exclude,family=gaussian(log))
summary(pricemodel2)
coef(pricemodel2)/log(10)
plot(real84_median_rent~year, data=SFok, type="p",log="y",
     xlab="year",ylab="Median rent (1983 dollars)",xlim=c(1975,2015))
lines(SFok$year, 10^(fitted(pricemodel)),col="purple")
lines(SFok$year, (fitted(pricemodel2)),col="forestgreen")
legend("bottomright",lty=1,col=c("purple","forestgreen"),
       legend=c("fit on log scale","fit on raw scale"),bty="n")
```

Still not quite the same, but close enough.

As Eric found, San Francisco real median asking prices for rentals have been going up about 2.5% per year since the 1950s,  and can be modelled quite accurately since 1975 using employment, real salaries and wages, and number of housing units.
